---
title: "Bikeshare Geographic Analysis"
output: html_notebook
---

```{r}
library(tidyverse)
library(here)
library(janitor)
library(lubridate)
library(sf)
library(tmap)
library(tidycensus)

bikeshare = read_csv(here("data_raw", "202209-capitalbikeshare-tripdata.zip")) %>%
  clean_names() %>% select(ride_id, start_lat, start_lng, end_lat, end_lng) %>%
  filter(!is.na(end_lat), !is.na(end_lng))

census=get_acs(geography = "tract",
                  variables=c("pop"="B01001_001"),
                  state="DC",geometry=TRUE,year=2018)
```

Count number of rides starting and ending in each census tract:
```{r}
census_wide = census %>% select(-moe) %>% spread(variable, estimate) %>% 
  select(GEOID, pop, geometry) %>% st_transform(4326)

bikeshare_start_pts = bikeshare %>% select(start_lng, start_lat) %>% 
  st_as_sf(coords=c("start_lng", "start_lat"), crs=st_crs(census_wide))
bike_census = census_wide %>% mutate(start_count = 
                                       lengths(st_intersects(., bikeshare_start_pts)))

bikeshare_end_pts = bikeshare %>% select(end_lng, end_lat) %>% 
  st_as_sf(coords=c("end_lng", "end_lat"), crs=st_crs(census_wide))
bike_census = bike_census %>% mutate(end_count = 
                                       lengths(st_intersects(., bikeshare_end_pts)))
```


Plot:
```{r}
tmap_mode("view")
tm_shape(bike_census) + tm_polygons(c("pop", "start_count", "end_count"), alpha=.5)
```

Normalized for population:
```{r}
# Filter out the one region specified because their bikeshare rate was so high that
# it skewed the visualization
bike_census_pop = bike_census %>% mutate(start_rate = start_count / pop,
                                         end_rate = end_count / pop) %>%
  filter(GEOID != 11001006202)
tm_shape(bike_census_pop) + tm_polygons(c("start_rate", "end_rate"), alpha=.5)
```

What other interesting questions are there relating to this data?
- How does location pattern vary with time of day?
- How does ridership change over the course of the year (tourism, etc.)?
- How does neighborhood age affect ridership?
    - Spatial age distribution in DC
- Do bikes get used up? In which areas?
- Is there a spatial pattern driven by members/casual riders?
- Is there a correlation between ethnicity and biking activity?
- Are metro stations a source of riders (last mile)?
- Do bike lanes affect biking activity?
- Does criminal activity near a location affect ridership patterns?
- Does socioeconomic standing of a neighborhood correlate with ridership?



Are metro stations a source of riders (last mile)?

look at tmap density plots and ggplot2 spatial visualizations