---
title: "Final Mile Original"
author: "Joseph LaPell, Tyler Yankee, Minsum Kim"
date: "2022-11-21"
output: html_document
---

This file is to be used for the original final mile analysis. Each person's own analysis will stem from this in new files.

Load packages
```{r}
library(tidyverse)
library(here)
library(janitor)
library(lubridate)
library(sf)
library(tmap)
library(geojsonsf)
library(matrixStats)
library(geosphere)
bikeshare = read_csv(here("data_raw", "202209-capitalbikeshare-tripdata.zip")) %>%
  clean_names()
metro_stations = read_csv(here("data_raw", "Metro_Stations_in_DC.csv")) %>%
  clean_names()
```

Define the Haversine distance function so we can compute distance between things
```{r}
# This function takes five arguments:
# 
# Longitude and latitude of the first location
# Longitude and latitude of the second location
# A parameter by which to round the responses

haversine <- function(long1, lat1, long2, lat2, round = 3) {
  # convert to radians
  long1 = long1 * pi / 180
  lat1  = lat1  * pi / 180
  long2 = long2 * pi / 180
  lat2  = lat2  * pi / 180
  
  R = 6371 # Earth mean radius in km
  
  a = sin((lat2 - lat1)/2)^2 + cos(lat1) * cos(lat2) * sin((long2 - long1)/2)^2
  d = R * 2 * asin(sqrt(a))
  
  return( round(d,round) ) # distance in km
}
```

Just doing a sample for a Monday, this is for September 5, 2022, the first Monday of the month.
```{r}
bikeshare1 = bikeshare %>% select(ride_id, slat = start_lat, slng = start_lng) %>% na.omit() 
metro_stations1 = metro_stations %>% select(end_station_name = name, lat = y, lng = x) %>% na.omit()
```

https://stackoverflow.com/questions/58831578/minimum-distance-between-lat-long-across-multiple-data-frames
Finds the closest station to bikes endpoint, adds station name and distance to the bikshare dataframe.

```{r}
#calculate the distance matrix
distmatrix<-distm(bikeshare1[, c(3, 2)], metro_stations1[,c(3, 2)])
#find closest column and get distance
closest<-apply(distmatrix, 1, which.min)
bikeshare1$close_station <- as.character(metro_stations1$end_station_name[closest])
bikeshare1$distance <- apply(distmatrix, 1, min)/1000
```

Counts the amount of bikes parked within 0.1524/500ft, close enough to consider bike being used for final mile.
```{r}
final_mile = bikeshare1 %>% filter(distance < 0.1524) 
bikeshare_end = bikeshare %>% select(ride_id, elat = end_lat, elng = end_lng)
new = left_join(final_mile, bikeshare_end, by = "ride_id")
```

```{r}
final_mile1 = new %>% mutate(start_to_end = haversine(slng, slat, elng, elat))
final_mile1 = final_mile1 %>% filter(start_to_end <= 1.60934 & start_to_end != 0) %>% select(ride_id, close_station, distance, start_to_end)
```

Make all of the data frames that we'll use for analysis.
```{r}
final_mile1_grouped = final_mile1 %>% group_by(close_station) %>% 
  summarise(count = n(), avg_distance_to_bikeshare = mean(distance), 
            avg_ride_distance = mean(start_to_end))
library(geojsonsf)
metro_stations_sf = metro_stations = geojson_sf(here("data_raw", "dc_metro_stations.geojson")) %>%
  clean_names() %>% select(close_station = name, metro_location = geometry)
final_mile1_geo = left_join(final_mile1_grouped, metro_stations_sf, by="close_station")
final_mile1_geo = st_as_sf(x = final_mile1_geo, crs = st_crs(metro_stations_sf))
```

* Analyze grouped data frame through distributions and comparing counts and average distances - Tyler.
* Visualizations: lines from start to end point for final mile rides, density distributions from above - Tyler.
* Join LODES job data with bikeshare end points that fall within final mile category - see if areas with more jobs also have more final mile rides - Joseph.
* Calculate proportion of rides within 500 ft of a station that are final mile rides - Minsum.
